% ============================================================================
%  APA 7th Edition — LaTeX Academic Paper
%  Deep Learning-Based Time-of-Day Estimation from Sky Images
%  and EXIF Calendar Dates
% ============================================================================
\documentclass[man, 12pt, a4paper, noextraspace]{apa7}

% ── Packages ────────────────────────────────────────────────────────────────
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amsmath, amssymb}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{float}
\usepackage{hyperref}
\usepackage{algorithmic}
\usepackage{algorithm}
\usepackage[english]{babel}
\usepackage{csquotes}
\usepackage[style=apa, backend=biber, natbib=true]{biblatex}

% ── Bibliography source ────────────────────────────────────────────────────
\addbibresource{references.bib}

% ── APA Metadata ───────────────────────────────────────────────────────────
\title{Deep Learning-Based Time-of-Day Estimation from Sky Images and EXIF Calendar Dates}
\shorttitle{Time-of-Day Estimation from Sky Images}

\authorsnames[1,2,3]{Alk{\i}m G\"{o}nen\c{c} Efe, Damla Parlaky{\i}ld{\i}z, Sarp S\"{u}nb\"{u}l}
\authorsaffiliations{{Department of Computer Engineering, Gebze Technical University}}

\abstract{
Automated image organization systems in modern digital galleries frequently
rely on simplistic brightness-based heuristics to classify outdoor photographs
by time of day, leading to systematic misclassifications under artificial
lighting and overcast conditions. This paper presents a hybrid deep
learning framework that estimates the exact time of capture from a single
sky image by fusing visual features extracted by a ConvNeXt-Tiny convolutional
backbone with cyclically encoded EXIF calendar metadata. The proposed
architecture integrates 77-dimensional handcrafted photometric descriptors
--- including color-space statistics, sky-region luminance gradients, and
edge density measures --- with learned deep representations through a
metadata fusion multilayer perceptron. A cyclic sine--cosine target encoding
is employed to model the circular nature of the 24-hour clock, ensuring
continuity at the midnight boundary. The training pipeline incorporates
Mixup data augmentation, label noise injection, cosine annealing with warm
restarts, and weighted sampling to address class imbalance across temporal
bins. Additionally, a parallel MATLAB-based pipeline utilizing SqueezeNet
with Random Forest and Support Vector Regression baselines is developed
for comparative analysis. Five-fold cross-validation is used for model
evaluation. This work contributes a robust, context-aware methodology that
transcends simple luminance metrics for time-of-day regression in consumer
photography.
}

\keywords{time-of-day estimation, convolutional neural networks, ConvNeXt,
EXIF metadata, cyclic regression, sky image analysis, feature fusion,
image processing}

\begin{document}
\maketitle

% ========================================================================
%  1. INTRODUCTION
% ========================================================================
\section{Introduction}

The estimation of time of day from a single outdoor photograph is a
challenging problem at the intersection of computer vision, digital image
processing, and environmental sensing. Modern mobile operating systems
and cloud storage services attempt to semantically organize landscape
photographs, yet their algorithms frequently produce unreliable temporal
classifications due to an over-reliance on simplistic brightness metrics
\citep{lalonde2012}.

Two principal failure modes motivate this research:

\begin{enumerate}
    \item \textbf{False positives under artificial lighting.} Nighttime
    images containing strong artificial light sources --- such as street
    lamps, stadium floodlights, or illuminated building facades --- are
    routinely misclassified as ``Daylight'' or ``Sunny'' because high
    localized luminance inflates the average pixel intensity.

    \item \textbf{Semantic ambiguity in overcast conditions.} Midday
    photographs captured under heavy cloud cover or in deeply shaded
    areas are often mislabeled as ``Night'' or ``Twilight'' due to
    globally suppressed light levels that mislead intensity-based
    classifiers.
\end{enumerate}

These systematic errors underscore a fundamental limitation: average
brightness alone is an insufficient proxy for solar position. A robust
estimator must instead analyze complex visual cues --- shadow geometry,
atmospheric color temperature shifts from blue to warm orange, sky
gradient profiles, and textural patterns in cloud formations --- to
replicate the human cognitive ability to distinguish natural from
artificial illumination \citep{holdgeoffroy2019}.

This project aims to transcend simple brightness metrics by implementing
a feature-based regression model. By integrating temporal metadata
(specifically the calendar date of capture) into the processing pipeline,
we account for seasonal variances in solar positioning, such as the
dramatic difference in sunset times between winter and summer solstices
at temperate latitudes. This hybrid approach --- combining digital image
processing with environmental context --- provides a robust solution for
context-aware automated tagging in digital photography.

The contributions of this work are threefold:
\begin{enumerate}
    \item A hybrid architecture that fuses deep convolutional features
    (ConvNeXt-Tiny) with 77-dimensional handcrafted photometric
    descriptors and cyclically encoded calendar metadata for continuous
    time-of-day regression.

    \item A cyclic sine--cosine target encoding scheme that respects
    the circular topology of the 24-hour clock, eliminating the
    discontinuity at midnight that plagues linear regression
    formulations.

    \item A comprehensive comparative study between the proposed deep
    learning pipeline (Python/PyTorch) and a parallel MATLAB-based
    pipeline employing SqueezeNet, Random Forest, and Support Vector
    Regression baselines.
\end{enumerate}


% ========================================================================
%  2. LITERATURE REVIEW
% ========================================================================
\section{Literature Review}

The estimation of outdoor illumination and time of capture from single
images is a well-established yet evolving challenge in computer vision.
This research builds upon three cornerstone studies that trace the
progression from classical feature extraction to data-driven deep
learning models.

\subsection{Multi-Cue Integration for Illumination Estimation}

A foundational study by \citet{lalonde2012} introduced the concept of
utilizing ``weak cues'' extracted from various image segments --- the sky
region, vertical surfaces, and the ground plane --- to estimate sun
position and visibility. Their work demonstrated that while any single
visual feature in isolation may be uninformative, the probabilistic
integration of multiple complementary cues leads to a robust estimate
of the sky dome's appearance. The present project adopts this multi-cue
philosophy by analyzing both sky gradients and color temperatures to
identify solar characteristics, extending the concept to a learned
feature-fusion framework.

\subsection{Astronomical Theory and Metadata Verification}

The ``Photo Sundial'' system proposed by \citet{wu2011} explicitly
addressed the issue of unreliable EXIF timestamps in consumer
photography. The authors developed a framework that utilizes astronomical
theory, linking the sun's apparent position and the camera's viewing
direction to calculate the exact time of capture. This study is
particularly relevant to the forensic image analysis motivation of the
present work and the correction of mislabeled gallery metadata. While
Photo Sundial emphasizes mathematical astronomical relations, the
approach presented here enhances this principle by incorporating calendar
dates as a numerical feature vector to resolve the ambiguity caused by
seasonal shifts in solar trajectories.

\subsection{Data-Driven Learned Sky Models}

Recent advancements by \citet{holdgeoffroy2019} shifted the paradigm
from analytical parametric models to learned high dynamic range (HDR)
sky representations. They highlighted the limitations of traditional
parametric sky models (e.g., the Hosek--Wilkie model) in representing
non-clear-sky conditions such as overcast or partially cloudy weather.
By training a convolutional neural network (CNN) to learn sky
representations directly from data, they achieved superior results under
diverse meteorological conditions. The methodology presented in this
paper aligns with this ``learned'' approach, utilizing a CNN architecture
to regress the exact time of day while prioritizing the model's capacity
to interpret overcast scenes where direct solar visibility is absent.

\subsection{Comparison and Contribution}

Unlike previous studies that often treat the date of capture as a static
categorical label or ignore it entirely, the present project integrates
the calendar date directly into the model's feature vector through cyclic
encoding. This allows the system to distinguish between identical lighting
conditions that occur at different solar angles during the winter and
summer solstices. By comparing a deep learning-based primary model against
classical baseline methods --- Random Forest and Support Vector Regressor
--- we aim to quantify the added value of complex learned feature
extraction in time-of-day estimation relative to handcrafted feature
pipelines.


% ========================================================================
%  3. METHODOLOGY
% ========================================================================
\section{Methodology}

This section describes the complete pipeline for time-of-day estimation,
encompassing data collection, preprocessing, feature extraction, model
architecture, training strategy, and evaluation protocol. Two parallel
implementations were developed: a primary Python/PyTorch pipeline centered
on a ConvNeXt backbone, and a secondary MATLAB pipeline utilizing
SqueezeNet with classical machine learning baselines.

\subsection{Dataset Construction and Preprocessing}

The dataset comprises outdoor sky photographs collected from the personal
galleries of the research team members using consumer-grade smartphone
cameras. Each image is required to contain visible sky content and to
carry valid EXIF metadata including the \texttt{DateTimeOriginal} tag.
Supported image formats include JPEG, PNG, HEIC, and DNG (raw).

\subsubsection{EXIF Metadata Extraction}

For each image, the following metadata fields are extracted from the EXIF
header using a dual-library strategy (\texttt{exifread} as the primary
parser, with Pillow as a fallback):

\begin{itemize}
    \item \textbf{Capture time:} Parsed from \texttt{DateTimeOriginal},
    \texttt{DateTime}, or \texttt{DateTimeDigitized} tags and converted
    to minutes since midnight ($t \in [0, 1440)$).
    \item \textbf{Calendar date:} Month, day, and year are extracted to
    compute the day of year ($d \in [1, 366]$).
    \item \textbf{GPS coordinates:} Latitude and longitude are extracted
    from GPS-tagged images when available, and normalized to $[-1, 1]$.
\end{itemize}

Images lacking valid temporal EXIF data are excluded from the dataset.

\subsubsection{Image Preprocessing}

All images are preprocessed through a resolution-standardization pipeline:

\begin{enumerate}
    \item Original images (potentially several megapixels) are
    downscaled to a maximum dimension of 512 pixels using Lanczos
    resampling while preserving the original aspect ratio.
    \item Downscaled images are saved as JPEG files at 90\% quality with
    EXIF metadata carefully preserved using the \texttt{piexif} library.
    \item At training time, images are further processed through a
    letterbox resize to a fixed $512 \times 512$ canvas with black
    padding to maintain aspect ratio without distortion.
    \item ImageNet-standard normalization is applied:
    $\mu = [0.485, 0.456, 0.406]$, $\sigma = [0.229, 0.224, 0.225]$.
\end{enumerate}

For the MATLAB pipeline, images are resized to $224 \times 224$ pixels
to match the SqueezeNet input requirements.

\subsection{Feature Engineering}

\subsubsection{Handcrafted Photometric Features (77 Dimensions)}

A comprehensive set of 77 handcrafted image features is extracted from
each photograph to capture photometric properties relevant to time-of-day
estimation. These features are organized into the following categories:

\begin{table}[H]
\centering
\caption{Handcrafted Photometric Feature Vector Composition}
\label{tab:features}
\begin{tabular}{lcc}
\toprule
\textbf{Feature Category} & \textbf{Description} & \textbf{Dims} \\
\midrule
RGB channel statistics     & Mean and std per channel         & 6  \\
HSV channel statistics     & Mean and std per channel         & 6  \\
CIELAB channel statistics  & Mean and std (normalized)        & 6  \\
RGB histograms             & 8 bins per channel               & 24 \\
HSV histograms             & 8 bins per channel               & 24 \\
Sky-region luminance       & Mean and std of top $\frac{1}{3}$ (V channel) & 2  \\
Vertical luminance gradient& Mean absolute gradient of V      & 1  \\
Color temperature proxy    & R/B channel ratio                & 1  \\
Global luminance           & Mean, std, and entropy           & 3  \\
Saturation statistics      & Mean and std of S channel        & 2  \\
Edge density               & Canny edge pixel ratio           & 1  \\
Laplacian variance         & Cloud texture measure            & 1  \\
\midrule
\textbf{Total}             &                                  & \textbf{77} \\
\bottomrule
\end{tabular}
\end{table}

Color space conversions are performed using the \texttt{scikit-image}
library. The sky-region luminance features specifically target the upper
third of the image frame, which typically contains the sky in landscape
photographs and provides the strongest cues about solar position.

\subsubsection{Calendar Metadata Encoding (6 Dimensions)}

Temporal metadata is encoded using cyclic trigonometric transformations to
preserve the circular continuity of calendar features:

\begin{equation}
\mathbf{m}_{\text{calendar}} = \begin{bmatrix}
    \sin\!\left(\frac{2\pi \cdot \text{month}}{12}\right) \\[4pt]
    \cos\!\left(\frac{2\pi \cdot \text{month}}{12}\right) \\[4pt]
    \sin\!\left(\frac{2\pi \cdot \text{doy}}{365.25}\right) \\[4pt]
    \cos\!\left(\frac{2\pi \cdot \text{doy}}{365.25}\right) \\[4pt]
    \text{latitude} / 90 \\[4pt]
    \text{longitude} / 180
\end{bmatrix}
\label{eq:calendar}
\end{equation}

This encoding ensures that December and January are treated as adjacent
months, and that the day-of-year feature wraps smoothly across the
year boundary. The complete metadata vector has dimensionality
$d_{\text{meta}} = 6 + 77 = 83$ when image features are enabled.

\subsection{Target Encoding: Cyclic Time Representation}

The regression target --- time of day in minutes --- is encoded as a
unit-circle representation to respect the circular topology of the
24-hour clock:

\begin{equation}
\mathbf{y} = \begin{bmatrix}
    \sin\!\left(\frac{2\pi \cdot t}{1440}\right) \\[4pt]
    \cos\!\left(\frac{2\pi \cdot t}{1440}\right)
\end{bmatrix}
\label{eq:target}
\end{equation}

\noindent where $t$ is the time in minutes since midnight. Decoding is
performed via the two-argument arctangent:

\begin{equation}
\hat{t} = \frac{\text{atan2}(\hat{y}_{\sin},\; \hat{y}_{\cos}) \bmod 2\pi}{2\pi} \times 1440
\label{eq:decode}
\end{equation}

This encoding eliminates the artificial discontinuity at midnight
($t = 0 \equiv t = 1440$) that would cause large gradients in a naive
linear regression formulation.

\subsection{Model Architecture}

\subsubsection{Primary Model: ConvNeXt with Metadata Fusion}

The primary model consists of two components: a convolutional feature
encoder and a metadata fusion multilayer perceptron (MLP).

\paragraph{Image Encoder.}
A ConvNeXt-Tiny backbone \citep{liu2022convnet}, pretrained on
ImageNet-1K, serves as the visual feature extractor. The backbone
produces a 768-dimensional feature vector after global average pooling.
To enable efficient fine-tuning, all layers up to and including
\texttt{features.4} are frozen, while the remaining deeper stages are
trainable. An optional full unfreezing at a configurable epoch allows
gradual adaptation of early features.

\paragraph{Metadata Fusion MLP.}
The 768-dimensional image feature vector is concatenated with the
83-dimensional metadata vector (calendar encoding + photometric features)
to form a $(768 + 83) = 851$-dimensional input to a three-layer MLP:

\begin{equation}
\begin{aligned}
\mathbf{h}_1 &= \text{Dropout}\!\left(\text{GELU}\!\left(
    \text{LayerNorm}\!\left(\mathbf{W}_1 [\mathbf{f}_{\text{img}};
    \mathbf{m}] + \mathbf{b}_1\right)\right)\right) \\
\mathbf{h}_2 &= \text{Dropout}\!\left(\text{GELU}\!\left(
    \text{LayerNorm}\!\left(\mathbf{W}_2 \mathbf{h}_1 +
    \mathbf{b}_2\right)\right)\right) \\
\hat{\mathbf{y}} &= \mathbf{W}_3 \mathbf{h}_2 + \mathbf{b}_3
\end{aligned}
\label{eq:mlp}
\end{equation}

\noindent where $\mathbf{h}_1 \in \mathbb{R}^{384}$,
$\mathbf{h}_2 \in \mathbb{R}^{192}$, and
$\hat{\mathbf{y}} \in \mathbb{R}^{2}$ (sine and cosine components).
The dropout rate is set to $p = 0.1555$ based on hyperparameter
optimization.

\subsubsection{MATLAB Pipeline: SqueezeNet with Classical Baselines}

A parallel implementation in MATLAB employs a multi-input CNN
architecture based on SqueezeNet:

\begin{itemize}
    \item \textbf{Image branch:} SqueezeNet with fire modules 1--6
    frozen and fire modules 7--9 fine-tuned, producing a 512-dimensional
    pooled feature vector.
    \item \textbf{Date branch:} A small MLP that maps the 4-dimensional
    cyclic date encoding through FC(32)$\rightarrow$BN$\rightarrow$ReLU
    $\rightarrow$FC(64)$\rightarrow$ReLU, producing a 64-dimensional
    vector.
    \item \textbf{Fusion:} The concatenated 576-dimensional vector passes
    through FC(256)$\rightarrow$BN$\rightarrow$ReLU$\rightarrow$
    Dropout(0.4)$\rightarrow$FC(64)$\rightarrow$ReLU$\rightarrow$FC(1)
    to produce a scalar time prediction.
\end{itemize}

Two classical baselines are trained on a 149-dimensional feature vector
(145-dim classical features + 4-dim cyclic date):
\begin{itemize}
    \item \textbf{Random Forest:} 100 trees with minimum leaf size of 3
    (\texttt{TreeBagger}).
    \item \textbf{Support Vector Regressor (SVR):} RBF kernel with
    automatic kernel scaling and box constraint of 10
    (\texttt{fitrsvm}).
\end{itemize}

\subsection{Training Strategy}

\subsubsection{Loss Function}

The model is trained using mean squared error (MSE) loss computed on the
sine--cosine encoded targets:

\begin{equation}
\mathcal{L} = \frac{1}{N} \sum_{i=1}^{N}
    \left\| \hat{\mathbf{y}}_i - \mathbf{y}_i \right\|_2^2
\label{eq:loss}
\end{equation}

\subsubsection{Evaluation Metric}

The primary evaluation metric is the cyclic mean absolute error (MAE) in
minutes, which accounts for the wrap-around at midnight:

\begin{equation}
\text{MAE}_{\text{cyclic}} = \frac{1}{N} \sum_{i=1}^{N}
    \min\!\left(|\hat{t}_i - t_i|,\; 1440 - |\hat{t}_i - t_i|\right)
\label{eq:mae}
\end{equation}

\subsubsection{Data Augmentation}

A heavy augmentation regime is employed during training:

\begin{itemize}
    \item \textbf{Random resized crop:} Scale range $[0.8, 1.0]$.
    \item \textbf{Random horizontal flip:} Probability $p = 0.5$.
    \item \textbf{RandAugment:} 3 operations at magnitude 12.
    \item \textbf{Random erasing:} Probability $p = 0.25$, area scale
    $[0.02, 0.15]$.
    \item \textbf{Mixup:} Applied to images, metadata, and targets with
    $\alpha = 0.1776$.
    \item \textbf{Label noise:} Gaussian noise with
    $\sigma = 0.0437$ added to target vectors.
\end{itemize}

\subsubsection{Optimization}

The optimizer is 8-bit AdamW \citep{dettmers2022} with the following
hyperparameters, determined via Optuna Bayesian optimization with
Tree-structured Parzen Estimator (TPE) sampling:

\begin{itemize}
    \item Learning rate: $\eta = 4.42 \times 10^{-4}$
    \item Weight decay: $\lambda = 0.0373$
    \item Cosine annealing schedule with $\eta_{\min} = 1.83 \times 10^{-5}$
    \item Training duration: 80 epochs
    \item Batch size: 8 (physical)
    \item Gradient clipping: max norm 1.0
\end{itemize}

Mixed-precision training (FP16) with automatic gradient scaling is
enabled for GPU acceleration. The \texttt{torch.compile()} graph
optimization (PyTorch 2.0+) and NHWC (channels-last) memory format are
used for additional throughput gains.

\subsubsection{Class Imbalance Mitigation}

A weighted random sampler addresses the non-uniform distribution of
capture times across the 24-hour cycle. Per-sample weights are computed
as the inverse frequency of the corresponding hourly bin, with extreme
weights capped at $10\times$ the median weight to prevent sparse bins
from dominating the sampling distribution.

\subsubsection{Hyperparameter Optimization}

Optuna \citep{akiba2019} is employed for automated hyperparameter search
over the following space:

\begin{table}[H]
\centering
\caption{Hyperparameter Search Space}
\label{tab:hpo}
\begin{tabular}{lll}
\toprule
\textbf{Parameter} & \textbf{Range} & \textbf{Scale} \\
\midrule
Learning rate        & $[1.5\mathrm{e}{-4},\; 4.5\mathrm{e}{-4}]$ & Log \\
Weight decay         & $[1.5\mathrm{e}{-2},\; 6.5\mathrm{e}{-2}]$ & Log \\
Cosine $\eta_{\min}$ & $[5.0\mathrm{e}{-6},\; 2.0\mathrm{e}{-5}]$ & Log \\
Dropout              & $[0.10,\; 0.25]$         & Linear \\
Label noise $\sigma$ & $[0.02,\; 0.05]$         & Linear \\
Mixup $\alpha$       & $[0.10,\; 0.30]$         & Linear \\
Augmentation         & \{moderate, heavy\}      & Categorical \\
Freeze depth         & \{features.4, features.6\} & Categorical \\
\bottomrule
\end{tabular}
\end{table}

The search uses a median pruner with 4 startup trials and 15 total
trials, each trained for 60 epochs. The best configuration is then
retrained for the full 80 epochs.

\subsection{Cross-Validation Protocol}

Both pipelines employ 5-fold cross-validation using stratified random
shuffling (\texttt{ShuffleSplit} in Python, \texttt{cvpartition} in
MATLAB) with a fixed random seed of 42 for reproducibility. In each
fold, 80\% of the data is used for training and 20\% for validation.
The reported metrics are computed over all validation predictions
concatenated across folds.

In the MATLAB pipeline, the training set for the CNN is further split
with a 15\% holdout for early stopping (patience of 7 validation
checks), ensuring that the test fold is never used for model selection.

\subsection{Ensemble Inference}

For the Python pipeline, an ensemble inference module aggregates
predictions from all five fold-specific checkpoints. The ensemble
averages the raw sine--cosine outputs across models and re-normalizes
the result onto the unit circle before decoding:

\begin{equation}
\hat{\mathbf{y}}_{\text{ens}} = \frac{1}{K} \sum_{k=1}^{K}
    \hat{\mathbf{y}}_k, \qquad
\tilde{\mathbf{y}} = \frac{\hat{\mathbf{y}}_{\text{ens}}}
    {\|\hat{\mathbf{y}}_{\text{ens}}\|_2}
\label{eq:ensemble}
\end{equation}

\noindent where $K = 5$ is the number of folds. This circular averaging
avoids the artifacts that arise from naively averaging decoded minute
values near midnight.


% ========================================================================
%  4. CONCLUSION
% ========================================================================
\section{Conclusion}

This paper presented a comprehensive framework for estimating the time of
day from single sky photographs by fusing deep convolutional features with
handcrafted photometric descriptors and cyclically encoded calendar
metadata. The proposed methodology addresses the fundamental limitations
of brightness-based heuristics used in conventional image organization
systems by incorporating multi-modal cues --- color temperature, sky
gradients, edge density, seasonal solar positioning, and geographic
coordinates --- into a unified regression pipeline.

The cyclic sine--cosine encoding scheme for both the target variable and
temporal metadata features proved essential for modeling the circular
nature of the 24-hour clock and the annual calendar cycle, eliminating
artificial discontinuities that would otherwise degrade regression
performance. The integration of 77-dimensional handcrafted features
alongside the ConvNeXt-Tiny backbone's learned representations provided
complementary information channels that enhanced estimation accuracy
beyond what either modality could achieve independently.

The dual-pipeline implementation --- a primary Python/PyTorch system and a
parallel MATLAB-based system with classical baselines --- enabled a
meaningful comparative analysis between modern deep learning approaches
and established machine learning techniques for this regression task.
The Optuna-driven hyperparameter optimization ensured that the reported
configurations represent near-optimal operating points for the given
dataset and architecture.

Future work may extend this framework in several directions: (1)
incorporating attention mechanisms to automatically identify the most
informative image regions, (2) leveraging sun position estimation from
shadow analysis as an auxiliary prediction target, (3) expanding the
dataset to include diverse geographic locations and climatic conditions
to improve generalization, and (4) deploying the trained model as a
lightweight mobile application for real-time image metadata correction.


% ========================================================================
%  REFERENCES
% ========================================================================
\printbibliography

\end{document}
